Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-05 manuscript/manuscript.tex
ported Jesse's comments from pdf to latex document

\title{Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages}

@@

\begin{abstract}
\noindent  
Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how site-specific functional constraints shift over time due to epistasis. 
\end{abstract}

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways that codon models account for purifying selection.}
(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
(B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
Most substitution models have stationary states that are uniform across sites.
However, substitution models can have site-specific stationary states.
In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
(C) Substitution models can incorporate neither, one, or both of these features.
Here we will use substitution models from the Goldman-Yang~\citep[GY94;][]{goldman1994codon,yang2000codon} and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed $\omega$ to represent all possible combinations.
}
\end{figure}

Proteins evolve under purifying selection to maintain their structure and function. 
This purifying selection is not homogenous across sites in a protein.
It is also not homogenous among the different amino acids at a given site.
For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.

Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
Codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average rate of fixation of nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
Here we will refer to M0 Goldman-Yang models simply as GY94 models. 
The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by~\citet{doud2016accurate}.
The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
For the GY94 model, the graphs are identical for all sites since this models does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
}
\end{figure}

Given a single branch, a substitution model transforms sequence identity into branch length.
Under a molecular-clock assumption, this branch length is proportional to time.
The transformation from sequence identity to branch length is trivial when the sequence identity is high.
For instance, when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
If the substitution model underestimates the asymptotic sequence identity then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

@@

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences measured by deep mutational scanning of an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree on the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
}
\end{figure}

@@

The models with a uniform stationary state underestimated the lengths of long branches separating the simulated sequences (\ref{fig:simulations}). 
The GY94 model estimated branches lengths which are only about 60\% of the true values for the longest branches. 
Accounting for site-to-site variation in $\omega$ did not fix the fundamental problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the longest branches.
However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models.
Therefore, models with uniform stationary states are fundamentally incapable of accurately estimating the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific stationary state and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree are optimized using the indicated ExpCM and GY94 models. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
The circle denotes the H1 clade and the triangle denotes the H3 clade. 
}
\end{figure}

\begin{table}[t!]
\caption{\label{tab:empirical_data}
{\bf Fitting of substitution models to the HA phylogenetic tree.}
The models fit here are the same ones in~\ref{fig:empirical_trees}. 
All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
All ExpCM's fit a stringency parameter $>1$.
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
       	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
	ExpCM (H1+H3 avg)  &  950 & -49227 & 0.15 & 1.78\\
	ExpCM + $\Gamma\omega$ (H1)  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
	ExpCM + $\Gamma\omega$ (H3) & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
	ExpCM (H1) & 2556 & -50030 &  0.13 & 1.22\\
	ExpCM (H3) &  3197 & -50350 & 0.12 & 1.45\\
	GY94 + $\Gamma\omega$  & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
	GY94 & 7625 & -52560  & 0.07 & -\\
      \end{tabular}
\end{table}

@@

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\citep{zhu2013hemagglutinin,hoffmann2016hemagglutinin,sun2013bat}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty probably lies in the length of the long branches separating the subtypes rather than in the tree topology itself.

There are two deep mutational scanning datasets for HA that estimated amino-acid preferences for all sites.
One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
\ref{suppfig:prefs_lee};][]{lee2018deep}.
Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
These three sets of HA amino-acid preferences define three different ExpCM's.

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each of these models with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all of the ExpCM's fit the actual data much better than the GY94 model.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ is generally fit to values $<1$ in phylogenetic studies~\citep{murrell2015gene} and our results follow this trend. 
\ref{tab:empirical_data} shows that most models were fit to very small $\omega$ values. 
However, the ExpCM's always fit $\omega$ values greater than the GY94 model and the ExpCM's with $\Gamma\omega$ all have one $\omega >1$. 
This increase in $\omega$ values makes sense due to the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
The ExpCM's more realistic null model gives it more power to detect diversifying selection, such as the \skhcomment{obvious} immune pressure on HA~\citep{bloom2017identification}.

\ref{fig:empirical_trees} shows that modeling purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state increases the estimated branch lengths.
But important, the effects from these two methods of modeling purifying selection are largely independent.
For every model (GY94 or any of the three ExpCM's), the $\Gamma\omega$ version estimates longer deep branches.
In addition, all ExpCM's estimate longer deep branches than the GY94 model, and all ExpCM+$\Gamma\omega$ models estimate longer deep branches than the GY94+$\Gamma\omega$ model.

>>>>

\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.}

@@

\begin{abstract}
\noindent  
Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
\skhcomment{I wonder if we should change this second sentence, because I'm not sure how many of the other cases for which there are actually other methods outside of viruses?Maybe it would be better to say: "For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times."Then the next sentence (as already written) talks about how for viruses there are other methods that show these are too recent.}
For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time due to epistasis. 
\end{abstract}

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways that codon models account for purifying selection.}
(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
(B) A substitution model's stationary state defines the expected sequence composition after a very long evolutionary time. 
Most substitution models have stationary states that are uniform across sites.
However, substitution models can have site-specific stationary states.
In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
(C) Substitution models can incorporate neither, one, or both of these features.
Here we will use substitution models from the Goldman-Yang~\citep[GY94;][]{goldman1994codon,yang2000codon} and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed $\omega$ to represent all possible combinations.
}
\end{figure}

Proteins evolve under purifying selection to maintain their structure and function. 
This purifying selection is not homogenous across sites in a protein.
It is also not homogenous among the different amino acids at a given site.
For instance, some protein sites strongly prefer hydrophobic amino acids, others are constrained to just one or a few amino acids, and yet others tolerate many amino acids.
In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.

Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average rate of fixation of nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
Here we will refer to M0 Goldman-Yang models simply as GY94 models. 
The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and usually better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
An alternative strategy of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} estimation is discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, \skhcomment{I realized this citation is wrong. For the fact that ExpCM and mutation-selection models can still incorporate omega, we should cite the relevant Rodrigue / Lartillot paper and Bloom (2017).The Haddox et al (2017) should be cited for the fact that omega can still be drawn from a gamma. At that point, we should also cite relevant Rodrigue / Lartillot paper for their approach to do this.} which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as experimentally measured by~\citet{doud2016accurate}.
The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
For the GY94 model, the graphs are identical for all sites since this model does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
\skhcomment{I still feel like some of the fonts here are too small. Can you make the small ones up the same size as the legend titles on the axes.}
}
\end{figure}

Given a single branch, a substitution model transforms sequence identity into branch length.
Under a molecular-clock assumption, this branch length is proportional to time.
The transformation from sequence identity to branch length is trivial when the sequence identity is high.
For instance, when there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some amino-acid asymptotic sequence identity.
For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
If the substitution model underestimates the asymptotic sequence identity then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

@@

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree on the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
}
\end{figure}

@@

The models with a uniform stationary state underestimated the lengths of long branches on the phylogenetic of the simulated sequences (\ref{fig:simulations}). 
The GY94 model estimated branches lengths that are only about 60\% of the true values for the longest branches. 
Accounting for site-to-site variation in $\omega$ did not fix the fundamental problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the longest branches.
However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models~\ref{fig:simulations}.
Therefore, models with uniform stationary states are fundamentally incapable of accurately estimating the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific stationary state and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree are optimized using the indicated ExpCM and GY94 models. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
The circle denotes the H1 clade and the triangle denotes the H3 clade. 
}
\end{figure}

\begin{table}[t!]
\caption{\label{tab:empirical_data}
{\bf Fitting of substitution models to the HA phylogenetic tree.}
The models fit here are the same ones in~\ref{fig:empirical_trees}. 
All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
All ExpCM's fit a stringency parameter $>1$.
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
       	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
	ExpCM (H1+H3 avg)  &  950 & -49227 & 0.15 & 1.78\\
	ExpCM + $\Gamma\omega$ (H1)  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
	ExpCM + $\Gamma\omega$ (H3) & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
	ExpCM (H1) & 2556 & -50030 &  0.13 & 1.22\\
	ExpCM (H3) &  3197 & -50350 & 0.12 & 1.45\\
	GY94 + $\Gamma\omega$  & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
	GY94 & 7625 & -52560  & 0.07 & -\\
      \end{tabular}
\end{table}

@@

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
We used \texttt{RAxML} with a nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.

There are two deep mutational scanning datasets for HA that estimated amino-acid preferences for all sites.
One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
\skhcomment{figure out how to put the lee paper first and then the supp figures}
The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
\ref{suppfig:prefs_lee};][]{lee2018deep}.
Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
These three sets of HA amino-acid preferences define three different ExpCM's.

We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification}\skhcomment{rodrigue MBE}, for amino-acid change, which occurs at some sites in HA due to immune selection.
